\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Fracture and failure of materials remain among the central problems of modern
mechanics and engineering. The ability to predict crack initiation and
propagation is essential for ensuring the reliability of structures operating
under complex mechanical and multiphysical loading conditions.

This book, \emph{Fracture Mechanics and Adjacent Fields: Insights from Ukrainian Research},
presents recent contributions of Ukrainian research schools in fracture
mechanics and related areas. The work builds upon strong traditions in
theoretical and applied mechanics and reflects contemporary developments in
analytical modelling, computational methods, and engineering applications.

The complete work comprises \textbf{43 chapters prepared by 120 authors}
representing leading institutes of the National Academy of Sciences of Ukraine.
To improve readability and thematic coherence, the material is presented in two
volumes.

\medskip
\textbf{The present volume (Volume~1)} focuses on the \emph{fundamental aspects}
of fracture mechanics and on \emph{interface phenomena}.

\medskip
\textbf{Part~I. Foundations of Fracture Mechanics} presents theoretical and
analytical approaches to crack problems, including stress intensity factors,
exact solutions, and modern computational techniques such as the boundary
element method. Special attention is given to fracture processes in
multiphysical media and generalized continua.

\medskip
\textbf{Part~II. Interface Cracks, Contact, and Adhesion} addresses fracture
mechanics of layered and composite systems. The chapters analyse interfacial
cracks under contact, friction, and adhesion, and consider dynamic and moving
load effects. These studies provide insight into fracture behaviour in bonded
structures and engineering interfaces.

\medskip
This volume establishes the theoretical and methodological basis for the
analysis of fracture processes, which is further developed and applied in the
second volume.

\medskip
The editors sincerely thank all authors for their valuable contributions and
their collaboration in preparing this work.

\bigskip
\noindent
Magdeburg, Kyiv, and Lviv \\
2026

\medskip
\noindent
\textit{Holm Altenbach}\\
\textit{Viacheslav Bogdanov}\\
\textit{Volodymyr Nazarenko}\\
\textit{Roman Kushnir}\\
\textit{Alexandr Grigorenko}\\
\textit{Mikhailo Selivanov}